---
title: "STAT3302FinalProject_Report"
author: "Dominick Yacono, Ben Gavie, Jack Kamnikar"
date: "2024-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#importing libraries
library(alr4)
library(tidyverse)
library(broom)

#importing datasets
game_data <- read.csv('NBAData/games.csv')
game_data_player_specific <- read.csv('NBAData/games_details.csv')
```

# Exploratory Data Analysis

## Removing Unnecessary Columns

To conduct our analysis of NBA data, we are utilizing two datasets. We are mainly interested in what skills are most effective toward affecting the odds of a home team winning. 

First, let's find which columns in the datasets will not be helpful in our analysis.

### games.csv file
```{r}
head(game_data,5)
```

The 'GAME_ID', 'HOME_TEAM_ID', 'VISITOR_TEAM_ID', and 'TEAM_ID_home' variables will be important to keep since they unique identify each game. Since we are looking at the skill/performance realted factors that influence victory, we can remove 'GAME_DATE_EST' and 'SEASON'. There also appear to be duplicate columns for the home team ID and away team ID, so we will remove these columns. 

```{r}
# Removing columns from game_data
columns_to_keep <- names(game_data)[!names(game_data) %in% c("GAME_DATE_EST", "SEASON", "TEAM_ID_home", "TEAM_ID_away")]
game_data <- subset(game_data, select = columns_to_keep)
```

### games_details.csv file
```{r}
head(game_data_player_specific,5)
```

We can drop 'TEAM_CITY', 'PLAYER_NAME', 'NICKNAME', 'COMMENT', and 'TEAM_ABBREVIATION' since these variables do not focus on player skill or performance during games.
We can drop 'START_POSITION' and 'MIN', because the information about the data on Kaggle informed us that only 62% of the rows have start position observations.

```{r}
# Removing columns from game_data
columns_to_keep <- names(game_data_player_specific)[!names(game_data_player_specific) %in% c("TEAM_CITY", "PLAYER_NAME", "NICKNAME", "COMMENT", "TEAM_ABBREVIATION", "START_POSITION", "MIN")]
game_data_player_specific <- subset(game_data_player_specific, select = columns_to_keep)

game_data_player_specific 
```

## Combining Into One Dataframe

Let's join the tables by GAME_ID

```{r}
# Removing columns from game_data
total_game_data <- merge(x = game_data_player_specific, y = game_data, by.x = "GAME_ID", by.y = "GAME_ID", all.x = FALSE, all.y = FALSE)
total_game_data <- total_game_data[order(total_game_data$GAME_ID), ] 

total_game_data
```

## Aggregating The Data