---
title: "STAT3302FinalProject_Report"
author: "Dominick Yacono, Ben Gavie, Jack Kamnikar"
date: "2024-04-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#importing libraries
library(alr4)
library(tidyverse)
library(broom)
library(ggplot2)

#importing datasets
game_data <- read.csv('NBAData/games.csv')
game_data_player_specific <- read.csv('NBAData/games_details.csv')
```

In this project, we are analyzing the factors that have an affect on the odds of a home team winning on home court in the NBA. 

To conduct our analysis of NBA data, we are utilizing two datasets—'games.csv' and 'game_details.csv'.  We are mainly interested in what skills are most effective toward affecting the odds of a home team winning. Because this dataset is so large and we have issues handling such large files on our systems, we will limit the data to the 2022 NBA season. 

```{R}
game_data <- game_data[game_data$SEASON == "2022", ]
```

# Exploratory Data Analysis

## Removing Unnecessary Columns

First, let's find which columns in the datasets will not be helpful in our analysis.

### games.csv file
```{r}
head(game_data,5)
```

The 'GAME_ID', 'HOME_TEAM_ID', and 'VISITOR_TEAM_ID' variables will be important to keep since they unique identify each game. Since we are looking at the skill/performance realted factors that influence victory, we can remove 'GAME_DATE_EST', 'SEASON', and 'GAME_STATUS_TEXT'. There also appear to be duplicate columns for the home team ID and away team ID, so we will remove these columns. 

```{r}
# Removing columns from game_data
columns_to_keep <- names(game_data)[!names(game_data) %in% c("GAME_DATE_EST", "SEASON", "GAME_STATUS_TEXT" ,"TEAM_ID_home", "TEAM_ID_away")]
game_data <- subset(game_data, select = columns_to_keep)
game_data
```

### games_details.csv file
```{r}
head(game_data_player_specific,5)
```

We can drop 'TEAM_CITY', 'PLAYER_NAME', 'NICKNAME', 'START_POSITION', 'MIN', and 'TEAM_ABBREVIATION' since these variables do not focus on player skill or performance during games. 
```{r}
# Removing columns from game_data
columns_to_keep <- names(game_data_player_specific)[!names(game_data_player_specific) %in% c("TEAM_CITY", "PLAYER_NAME", "NICKNAME", "TEAM_ABBREVIATION", "START_POSITION", "MIN")]
game_data_player_specific <- subset(game_data_player_specific, select = columns_to_keep)

game_data_player_specific 
```

We will remove 'COMMENT' as well, but first we must filter the data to not include players with comment of "DNP - Coach's Decision".  Players with this comments did not play in the game, so therefore they are not useful in our analysis.

```{r}
#Filtering the game_data_player_specific
game_data_player_specific_filtered <- game_data_player_specific[!game_data_player_specific$COMMENT != "", ]

#Removing "COMMENT" column
columns_to_keep <- names(game_data_player_specific_filtered)[!names(game_data_player_specific_filtered) %in% c("COMMENT")]
game_data_player_specific_filtered <- subset(game_data_player_specific_filtered, select = columns_to_keep)
game_data_player_specific_filtered

game_data_player_specific_filtered
```

## Combining Into One Dataframe

Let's join the tables by GAME_ID

```{r}
# Removing columns from game_data
total_game_data <- merge(x = game_data_player_specific_filtered, y = game_data, by.x = "GAME_ID", by.y = "GAME_ID", all.x = FALSE, all.y = FALSE)
total_game_data <- total_game_data[order(total_game_data$GAME_ID), ] 

total_game_data
```

Some columns between 'games.csv' and 'gamesdata.csv' are shared and cover the same attributes, so let' remove 'FCM', 'FGA', 'FG_PCT', 'FG3M', 'FG3A', 'FG3_PCT', 'FTM', 'FTA', 'FT_PCT', 'REB', and 'AST'. 



## Aggregating The Data

Now that the data has been combined, we must aggregate the player specific data. This will allow us to analyze how the team performed as a whole for each game.

First, we can see that we have 'FGM' (Field goals made) and 'FGA' (Field goals attempted). We can make an aggregated variables out of this. Let's make a variable 

```{R}
# Removing columns from total_game_data
columns_to_keep <- names(total_game_data)[!names(total_game_data) %in% c('FGM', 'FGA', 'FG_PCT', 'FG3M', 'FG3A', 'FG3_PCT', 'FTM', 'FTA', 'FT_PCT', 'REB', 'AST')]
total_game_data <- subset(total_game_data, select = columns_to_keep)

total_game_data
```

We will aggregate the data for just home teams, since this study is examining them.

#### Aggregating The Data For Home Teams
```{R}
# Only obtain players that played for the home team
home_team_stats_aggregated <- filter(total_game_data, TEAM_ID == HOME_TEAM_ID)
home_team_stats_aggregated

#Group by game_id and make aggregate data
home_team_stats_aggregated <- home_team_stats_aggregated %>% group_by(GAME_ID, HOME_TEAM_ID, VISITOR_TEAM_ID, PTS_home, FG_PCT_home, FT_PCT_home, FG3_PCT_home, AST_home, REB_home, HOME_TEAM_WINS) %>%
  summarize(OREB_home = sum(OREB), DREB_home = sum(DREB), STL_home = sum(STL), 
            BLK_home = sum(BLK), TO_home = sum(TO), PF_home = sum(PF), avg_PTS_per_player_home = mean(PTS), avg_PLUS_MINUS_per_player_home = mean(PLUS_MINUS)) 

#Remove GAME_ID, HOME_TEAM_ID, and VISITOR_TEAM_ID now since we are done bringing the data together 

columns_to_keep <- names(home_team_stats_aggregated)[!names(home_team_stats_aggregated) %in% c("GAME_ID","TEAM_ID", "HOME_TEAM_ID", "VISITOR_TEAM_ID")]
home_team_stats_aggregated <- subset(home_team_stats_aggregated, select = columns_to_keep)

home_team_stats_aggregated 
```

## Univariate Analysis

### Analyzing Central Tendency
```{R}
columns_to_keep <- names(home_team_stats_aggregated)[!names(home_team_stats_aggregated) %in% c("HOME_TEAM_WINS")]
home_team_stats_aggregated_no_response <- subset(home_team_stats_aggregated, select = columns_to_keep)

summary(home_team_stats_aggregated_no_response)
```

```{R}
column_names <- names(home_team_stats_aggregated_no_response)

for (col in column_names) {
  boxplot(home_team_stats_aggregated_no_response[, col], main = col, xlab = col, ylab = "", notch = FALSE)
}
```

### Analyzing Spread

```{R}
column_names <- names(home_team_stats_aggregated_no_response) 

density_plots <- lapply(column_names, function(col) {
  ggplot(home_team_stats_aggregated_no_response, aes_string(x = col)) +
    stat_density(aes(), geom = "area", alpha = 0.5) +  
    labs(title = col, x = col, y = "Density") +
    theme_bw()
})

density_plots

```

### Q-Q Plots and Normality

```{R}
qqnorm(home_team_stats_aggregated_no_response$PTS_home, main = "QQplot of PTS_Home")
qqline(home_team_stats_aggregated_no_response$PTS_home)

qqnorm(home_team_stats_aggregated_no_response$FG_PCT_home, main = "QQplot of FG_PCT_Home")
qqline(home_team_stats_aggregated_no_response$FG_PCT_home)

qqnorm(home_team_stats_aggregated_no_response$FT_PCT_home, main = "QQplot of FT_PCT_Home")
qqline(home_team_stats_aggregated_no_response$FT_PCT_home)

qqnorm(home_team_stats_aggregated_no_response$FG3_PCT_home, main = "QQplot of FG3_PCT_Home")
qqline(home_team_stats_aggregated_no_response$FG3_PCT_home)

qqnorm(home_team_stats_aggregated_no_response$AST_home, main = "QQplot of AST_Home")
qqline(home_team_stats_aggregated_no_response$AST_home)

qqnorm(home_team_stats_aggregated_no_response$REB_home, main = "QQplot of REB_Home")
qqline(home_team_stats_aggregated_no_response$REB_home)

qqnorm(home_team_stats_aggregated_no_response$OREB_home, main = "QQplot of OREB_Home")
qqline(home_team_stats_aggregated_no_response$OREB_home)

qqnorm(home_team_stats_aggregated_no_response$DREB_home, main = "QQplot of DREB_Home")
qqline(home_team_stats_aggregated_no_response$DREB_home)

qqnorm(home_team_stats_aggregated_no_response$STL_home, main = "QQplot of STL_Home")
qqline(home_team_stats_aggregated_no_response$STL_home)
  
qqnorm(home_team_stats_aggregated_no_response$BLK_home, main = "QQplot of BLK_Home")
qqline(home_team_stats_aggregated_no_response$BLK_home)

qqnorm(home_team_stats_aggregated_no_response$TO_home, main = "QQplot of TO_Home")
qqline(home_team_stats_aggregated_no_response$TO_home)

qqnorm(home_team_stats_aggregated_no_response$PF_home, main = "QQplot of PF_Home")
qqline(home_team_stats_aggregated_no_response$PF_home)

qqnorm(home_team_stats_aggregated_no_response$avg_PTS_per_player_home, main = "QQplot of Avg_PTS_per_player_Home")
qqline(home_team_stats_aggregated_no_response$avg_PTS_per_player_home)

qqnorm(home_team_stats_aggregated_no_response$avg_PLUS_MINUS_per_player_home, main = "QQplot of Avg_PLUS_MINUS_per_player_Home")
qqline(home_team_stats_aggregated_no_response$avg_PLUS_MINUS_per_player_home)
```
  
For the most part, all our variables appear to be somewhat normal. There is no need to do transformations.  

## Multivariate Analysis

### Pairplots



  
```